---
title: "R Notebook"
output: html_notebook
---

## 1. Introduction to Data Cleaning in R (30 minutes)


### a. Why Do We Need Data Cleaning? 
* Explanation: Data collected from various sources is often messy and needs to be cleaned to ensure accurate analysis. This includes handling missing values, correcting data types, removing duplicates, etc.
* Key Point: Data cleaning is the foundation of any reliable analysis.

### b. What Are Missing Values? 
* Explanation: Missing values occur when no data is available for some variables. This can cause issues in analysis.

```{r}
library(tidyverse)

# Load dataset with missing values
data("airquality")  # Pre-installed dataset in R
head(airquality)  # View first few rows of the dataset
glimpse(airquality)
str(airquality)
```

### c. Checking Missing Values 
* Explanation: Before cleaning, it's important to check how much of the data is missing.
```{r}
# Check for missing values in each column
colSums(is.na(airquality))

# Checking missing values for the whole data
table(is.na(airquality))

# Checking missing values for the whole data
summary(airquality)


```

### d. Putting Data in the Correct Format 
* Explanation: Variables may need to be converted into proper data types such as factors, numeric, etc.
```{r}
# Convert 'Month' column to a factor
airquality$Month <- as.factor(airquality$Month)
str(airquality)  # Check structure of the dataset

data("starwars")
starwars$gender <- factor(starwars$gender,
                          levels = c("masculine", "feminine"))

```

### e. Checking for Outliers 
* Explanation: Outliers can skew the results of an analysis and might need to be handled
```{r}
# Simple visualization to spot outliers
boxplot(airquality$Ozone, main = "Ozone Levels Boxplot")
```
The data two
### f. Handling Missing Values
* Explanation: There are different ways to handle missing values:
* Remove rows with missing values.
* Impute missing values (fill in with mean, median, or a constant).
```{r}
# Option 1: Remove rows with missing values
airquality_clean <- na.omit(airquality)

# Option 2: Impute missing values with the mean
airquality$Ozone[is.na(airquality$Ozone)] <- mean(airquality$Ozone, na.rm = TRUE)
head(airquality)


airquality$Ozone <- ifelse(is.na(airquality$Ozone), 
                           median(airquality$Ozone, na.rm = TRUE),
                           airquality$Ozone)
```

### g. Handling Duplicates 
* Explanation: Duplicate entries can distort analysis and need to be removed.

```{r}
# Check for duplicates
duplicated_rows <- duplicated(airquality)
sum(duplicated_rows)

# Remove duplicates
airquality_clean <- airquality[!duplicated(airquality), ]


# Or
airquality_clean <- airquality %>% 
  distinct()
```

### h. Cleaning Column Names
* Explanation: Column names should be clear and consistent.
```{r}
# Use the janitor package to clean column names
library(janitor)
airquality_clean <- clean_names(airquality)
head(airquality_clean)

```
########################-------- PRACTICE --------------------######################




## Data Manipulation
### 2. Introduction to Data Manipulation in R
Selecting Variables in tidyverse
```{r}
data = airquality %>% 
  select(Ozone, ,Solar.R, Wind)
```


Renaming variables
```{r}
# Rename columns in airquality dataset
airquality_clean <- airquality %>%
  rename(Temperature = temp, WindSpeed = wind)

head(airquality_clean)

```

Relocating variables
```{r}
# Reorder columns using relocate()

airquality_clean %>%
  relocate(Day, Ozone)
```

### Filtering Data
```{r}
# Filter data for specific conditions
high_temp <- airquality %>% filter(Temp > 80)  # Filter for temperatures greater than 80
head(high_temp)

data("starwars")
# Filtering where species were Human and Droid

filtered <- starwars %>% 
  filter(species %in% c("Human", "Droid"))

# Filtering using subset
subset(starwars, species %in% c("Human", "Droid"))

```

### Converting Data from Long to Wide Format 
* Explanation: Data can exist in long or wide format, and you may need to switch between the two.
```{r}
long_airquality <- pivot_longer(airquality, cols = c(Ozone, Temp, Wind), 
                                names_to = "Variable", values_to = "Value")
head(long_airquality)

# Pivot wider example (from long to wide format)
wide_airquality <- pivot_wider(long_airquality, names_from = "Variable", values_from = "Value")
head(wide_airquality)

```



### d. Aggregating Data (10 minutes)
* Explanation: Aggregating means summarizing data by groups, such as calculating the mean for different months.
```{r}
# Summarize data by Month (calculate the average for each variable)
airquality_summary <- airquality %>%
  group_by(Month) %>%
  summarize(mean_ozone = mean(Ozone, na.rm = TRUE),
            mean_temp = mean(Temperature, na.rm = TRUE),
            mean_wind = mean(WindSpeed, na.rm = TRUE))
airquality_summary

```


### Additional Manipulation Techniques
* Explanation: Other useful techniques include sorting and mutating (creating new columns).
```{r}
# Sorting data by temperature
sorted_data <- airquality_clean %>% arrange(desc(Temperature))

# Creating a new column
airquality_clean <- airquality_clean %>% mutate(Ozone_Level = ifelse(Ozone > 100, "High", "Low"))

```




