---
title: "R Notebook"
output: html_notebook
---
## Data Visualizations in R

### 1. Introduction to Data Visualizations
Purpose of data visualizations: to summarize, explore, and present data in a visually intuitive way.
Types of visualizations: histograms, boxplots, pie charts, bar charts, scatter plots.
We will use the mtcars dataset as our base for all demonstrations.


### Visualizations in Base R
#### a. Histogram: Distribution of Miles per Gallon (MPG)
```{r}
hist(mtcars$mpg, col = "blue", main = "Histogram of MPG", xlab = "Miles per Gallon", ylab = "Frequency")

```
Interpretation:
Shows the distribution of mpg. Most cars fall between 15 and 20 MPG.

#### b. Boxplot: MPG by Number of Cylinders
```{r}
boxplot(mpg ~ cyl, data = mtcars, col = "lightblue", main = "MPG by Number of Cylinders", xlab = "Cylinders", ylab = "Miles per Gallon")

```
Interpretation:

Higher mpg for 4-cylinder cars compared to 6- and 8-cylinder cars.

### c. Pie Chart: Proportion of Cars by Number of Cylinders
```{r}
cylinders_count <- table(mtcars$cyl)
pie(cylinders_count, main = "Proportion of Cars by Number of Cylinders", col = c("lightblue", "lightgreen", "lightpink"))

```
### d. Bar Chart: Frequency of Cars by Number of Gears
```{r}
gear_count <- table(mtcars$gear)
barplot(gear_count, col = "blue", main = "Frequency of Cars by Number of Gears", xlab = "Number of Gears", ylab = "Count")

```
### e. Scatter Plot: Relationship between Weight and MPG
```{r}
plot(mtcars$wt, mtcars$mpg, main = "Scatter Plot of Weight vs MPG", xlab = "Weight (1000 lbs)", ylab = "Miles per Gallon", col = "blue")

```
Negative correlation between weight and MPG. Heavier cars have lower MPG.


### 3. Visualizations Using ggplot2

```{r}
library(ggplot2)
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Histogram of MPG", x = "Miles per Gallon", y = "Frequency") +
  theme_minimal()


```
### b. Boxplot: MPG by Number of Cylinders
```{r}
ggplot(mtcars, aes(x = cyl, y = mpg, fill = factor(cyl))) +
  geom_boxplot() +
  labs(title = "MPG by Number of Cylinders", x = "Cylinders", y = "Miles per Gallon") +
  theme_minimal()

```
### d. Bar Chart: Frequency of Cars by Number of Gears
```{r}
ggplot(mtcars, aes(x = factor(gear), fill = factor(gear))) +
  geom_bar() +
  labs(title = "Frequency of Cars by Number of Gears", x = "Number of Gears", y = "Count") +
  theme_minimal()

```
### e. Scatter Plot: Relationship between Weight and MPG
```{r}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "blue", size = 5) +
  labs(title = "Scatter Plot of Weight vs MPG", x = "Weight (1000 lbs)", y = "Miles per Gallon") +
  theme_minimal()

```
```{r}



```

### Hands-on Exercises
Exercise 1: Create a histogram and boxplot using the iris dataset to explore sepal length.
Exercise 2: Use ggplot2 to create a scatter plot of horsepower vs miles per gallon in the mtcars dataset and color points based on transmission type.