---
title: "Matrices"
output:
  html_notebook: default
---

Matrices.
What is a matrix?
When a vector is introduced with row and column i.e. a dimension attribute, it becomes a matrix. A matrix is represented by set of rows and columns. It is a 2 dimensional data structure. 


Creating a matrix in R.
```{r}
# Creating a matrix. Let’s create a matrix of 3 rows and 2 columns
my_matrix = matrix(1:9, nrow=3, ncol=3)

# matrix() this function is used to create a matrix in R
# 1:9 creates a vector with the numbers from 1 to 9
# nrow = 3, Clarifies that the matrix should have 3 rows
# ncol = 3, Clarifies that the matrix should have 3 columns
```

Accessing Elements in a matrix
```{r}
# Accessing elements in a matrix
# So to accessing the elements we need to specify the row and column of interest
# Now let us access the first row and the third column
my_matrix[1, 3]
# The value 1 specifies the row we want 
# The value 3 specifies the column we want 

my_matrix[1, ]
# So in this example we just want to access only the first row of the matrix.

my_matrix[, 2] #or
my_matrix[2]
# Here we are accessing the second column of the matrix 
```
Checking the dimensions, attributes and class of a matrix
```{r}
# Checking the dimensions of a matrix (the number of rows and columns of a matrix)
dim(my_matrix)

# Checking the attributes of a matrx (basically the  dimensions)
attributes(my_matrix)

# Checking the class of a matrix
class(my_matrix)
```

Converting Vectors to Matrix
```{r}
# Creating an Age vector the converting it to matrix
age = c(23, 44, 15, 12, 31, 16)

# Converting a matrix to a vector by specifying the dimensions of the matrix
dim(age) <- c(2,3)

# Checking the class to see if it is a matrix
class(age)
```

Mathematical Operations in Matrices
```{r}
## Mathematical oparations in matrices
matrix_b = matrix(10:18, nrow = 3)

# Summing of matrices
my_matrix + matrix_b

# Multiplication of matrices
my_matrix %*% matrix_b

# Determinant of matrices
det(my_matrix)

# Transpose of a matrix
t(my_matrix)

# Inverse of matrix
# Let us create a new matrix
a1 <- c(3, 2, 5) 
a2 <- c(2, 3, 2) 
a3 <- c(5, 2, 4) 

new_matrix = matrix(c(a1, a2, a3), nrow = 3)  
solve(new_matrix)
```



Data Frames
What is a data frame?
This is the most commonly used member of data types family. It is used to store tabular data. It is different from matrix. In a matrix, every element must have same class. But, in a data frame, you can put list of vectors containing different classes. 
```{r}
# Creating a dataframe
df = data.frame(name = c("ash","jane","paul","mark"), score = c(67,56,87,91))

# View a data frame
df
print(df)
View(df)

# Checking the dimensions , structure and attributes of a data frame
dim(df)
attributes(df)
str(df)

```

More on Data Frames
```{r}
# Creating a simple data frame with students' names, ages, and marks
names = c("Brayo", "Gaturu", "Oliver", "Victor", "Elli")  # A vector of student names
Age = c(21, 80, 24, 20, 18)  # A vector of ages
Marks = c('A', 'B', 'C', 'D', 'A')  # A vector of marks

# Creating a data frame from the vectors
data = data.frame(names, Age, Marks)


# Creating another data frame in a single step, same as the one above
df = data.frame(names = c("Brayo", "Gaturu", "Oliver", "Victor", "Elli"),
                Age = c(21, 80, 24, 20, 18),
                Marks = c('A', 'B', 'C', 'D', 'A'))


```

Accessing specific columns and elements in the data frame
```{r}

data[1, 2]  # Retrieves the value in the 1st row and 2nd column (age of the 1st student)

# Accessing a range of rows using index numbers
data[1:3]  # Retrieves the first 3 rows of the data frame

# Accessing the first 2 rows of the data frame
data[1:2]

# Accessing columns using the `$` operator
data$Age  # Retrieves the 'Age' column
data$Marks  # Retrieves the 'Marks' column
data$names  # Retrieves the 'names' column

# Adding a new column 'Gender' to the data frame
data$Gender = c("Male", "Female", "Male", "Female", "Male")
```

Data frames operations
```{r}
# Modifying values in specific cells in the data frame
data[1, 4:5] = 'NA'  # Assigning 'NA' to the 1st row and 4th, 5th columns (Gender and Marks)


# Removing the 5th column from the data frame
data = data[, -5]  # The '-5' means removing the 5th column
data  # Displaying the updated data frame

# Accessing a specific column by its name using square brackets
data[,"names"]  # Retrieves the 'names' column

```

Working with actual Data frames
```{r}
# Loading built-in data sets in R
data = datasets::women  # Loading the 'women' data set
class(data)  # Checking the class (type) of the data set
str(data)  # Checking the structure of the data set
head(data)  # Displaying the first 6 rows of the data set
tail(data)  # Displaying the last 6 rows of the data set

# Working with another built-in data set 'ability.cov'
data_2 = datasets::ability.cov  # Loading the 'ability.cov' data set
class(data_2)  # Checking the class of the data set
head(data_2)  # Displaying the first few rows of the data set
View(data_2)  # Viewing the data set in RStudio
```


More
```{r}

# Converting the 'ability.cov' dataset to a data frame
data_2 = as.data.frame(data_2)
data_2  # Displaying the converted data frame

# Working with the Titanic dataset
DATA = datasets::Titanic  # Loading the 'Titanic' dataset
class(DATA)  # Checking the class of the dataset

# Converting the Titanic dataset to a data frame
DATA = as.data.frame(DATA)
DATA  # Displaying the data frame

# Accessing specific columns in a data frame
DATASET$Survived  # Accessing the 'Survived' column
DATASET$Age  # Accessing the 'Age' column
DATASET[, 'Age']  # Accessing the 'Age' column using square brackets

```

